---
title: "Expression Data Integration Pipeline"
subtitle: "Multi-dataset integration with batch correction, PCA, differential expression, and network clustering"
author: "Expression Integration Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 10,
  fig.height = 8
)
```

# Overview

This pipeline integrates gene expression data from multiple studies/datasets and performs:

1. **Data Merging** - Combine expression matrices by common genes
2. **Batch Correction** - Remove technical batch effects using ComBat
3. **PCA Analysis** - Visualize data structure and sample relationships
4. **Differential Expression** - Identify differentially expressed genes using limma
5. **Network Clustering** - Build protein-protein interaction networks and detect communities using fastgreedy algorithm
6. **Functional Enrichment** - Perform pathway enrichment analysis cluster-by-cluster

---

# 1. Setup

## Load Libraries

```{r load_libraries, message=FALSE}
# Load pipeline modules
source("../R/config.R")
source("../R/utils.R")
source("../R/01_data_merging.R")
source("../R/02_batch_correction.R")
source("../R/03_pca_analysis.R")
source("../R/04_differential_expression.R")
source("../R/05_network_clustering.R")
source("../R/06_enrichment_analysis.R")
source("../R/07_visualization.R")

# Load required packages
library(limma)
library(sva)
library(factoextra)
library(ggplot2)
library(igraph)
```

## Load Configuration

```{r load_config}
# Load configuration from YAML file
config <- load_config("../config/config.yaml")

# Validate and print configuration
validate_config(config)
print_config(config)

# Create output directories
output_dirs <- create_output_dirs(config$paths$output)
```

---

# 2. Data Merging

Merge expression data from multiple datasets, keeping only genes common to all studies.

```{r merge_data}
log_msg("Starting data merging...")

# Merge expression matrices
merged_exprs <- merge_expression_data(
  mapped_path = config$paths$mapped_data,
  pattern = config$merging$pattern
)

# Load phenodata
pdata <- read.csv(config$paths$phenodata, stringsAsFactors = FALSE)

# Align phenodata with expression data
pdata_aligned <- align_phenodata(
  mrgd = merged_exprs,
  pdata = pdata,
  sample_col = config$merging$sample_column
)

# Display dimensions
cat("\nMerged expression data:\n")
cat("  Genes:", nrow(merged_exprs), "\n")
cat("  Samples:", ncol(merged_exprs), "\n")
cat("  Phenodata samples:", nrow(pdata_aligned), "\n")
```

## Apply Filters (Optional)

```{r filter_data}
# Apply sample filters if defined in config
if (!is.null(config$filtering) && length(config$filtering) > 0) {

  log_msg("Applying sample filters...")

  filtered_data <- filter_samples(
    mrgd = merged_exprs,
    pdata = pdata_aligned,
    filters = config$filtering
  )

  merged_exprs <- filtered_data$exprs
  pdata_aligned <- filtered_data$pdata

} else {
  log_msg("No filters defined, using all samples")
}

# Validate input data
validate_input(merged_exprs, pdata_aligned)
```

---

# 3. Batch Correction

Remove technical batch effects using ComBat while preserving biological variation.

```{r prepare_batch_correction}
log_msg("Preparing for batch correction...")

# Add trimester variable if needed
if ("trim_term" %in% all.vars(as.formula(config$batch_correction$model_formula))) {
  pdata_aligned <- add_trimester_variable(pdata_aligned)
}

# Check batch distribution
cat("\nBatch distribution:\n")
print(table(pdata_aligned[[config$batch_correction$batch_column]]))
```

```{r batch_correction}
log_msg("Applying ComBat batch correction...")

# Apply batch correction
exprs_corrected <- apply_combat_correction(
  exprs = merged_exprs,
  pdata = pdata_aligned,
  batch_col = config$batch_correction$batch_column,
  mod_formula = as.formula(config$batch_correction$model_formula),
  par_prior = config$batch_correction$par_prior,
  prior_plots = config$batch_correction$prior_plots
)

# Save corrected data
save_data(exprs_corrected,
          file.path(output_dirs$base, "exprs_corrected.tsv"),
          format = "tsv")
save_data(pdata_aligned,
          file.path(output_dirs$base, "pdata_aligned.tsv"),
          format = "tsv")
```

---

# 4. PCA Analysis

Perform principal component analysis to visualize data structure.

```{r pca}
log_msg("Performing PCA...")

# Perform PCA
pca <- perform_pca(exprs_corrected,
                   center = config$pca$center,
                   scale = config$pca$scale)

# Create scree plot
plot_pca_scree(pca,
               output_dir = output_dirs$plots,
               n_pcs = config$pca$n_pcs)
```

## PCA Plots by Metadata

```{r pca_plots}
log_msg("Creating PCA plots...")

# Plot PCA colored by different variables
plot_pca_metadata(pca,
                  pdata = pdata_aligned,
                  variables = config$pca$color_by,
                  output_dir = output_dirs$plots,
                  axes = c(1, 2))

# Create biplot if specified
if (!is.null(config$pca$biplot_genes) && config$pca$biplot_genes > 0) {
  plot_pca_biplot(pca,
                  pdata = pdata_aligned,
                  color_by = config$pca$color_by[1],
                  n_genes = config$pca$biplot_genes,
                  output_dir = output_dirs$plots)
}
```

---

# 5. Differential Expression Analysis

Identify differentially expressed genes using limma.

```{r differential_expression}
log_msg("Performing differential expression analysis...")

# Perform differential expression
de_results <- differential_expression(
  exprs = exprs_corrected,
  pdata = pdata_aligned,
  design_formula = as.formula(config$differential_expression$design_formula),
  contrasts = config$differential_expression$contrasts,
  method = config$differential_expression$decide_method
)

# Extract and annotate results
difexp <- extract_difexp_results(
  efit = de_results$efit,
  results = de_results$results,
  exprs = exprs_corrected,
  pdata = pdata_aligned
)

# Filter by thresholds
difexp_filtered <- filter_difexp(
  difexp,
  logfc_threshold = config$differential_expression$logfc_threshold,
  pvalue_threshold = config$differential_expression$pvalue_threshold
)

# Save results
save_data(difexp,
          file.path(output_dirs$difexp, "difexp_all.csv"))
save_data(difexp_filtered,
          file.path(output_dirs$difexp, "difexp_filtered.csv"))

cat("\nDifferential expression summary:\n")
cat("  Total genes tested:", nrow(difexp), "\n")
cat("  Significant genes:", nrow(difexp_filtered), "\n")
```

## Volcano Plot

```{r volcano_plot}
log_msg("Creating volcano plot...")

plot_volcano(
  difexp,
  output_file = file.path(output_dirs$plots, "volcano_plot.png"),
  logfc_threshold = config$differential_expression$logfc_threshold,
  pval_threshold = config$differential_expression$pvalue_threshold,
  label_top = config$visualization$volcano_top_genes
)
```

## Heatmap

```{r heatmap}
log_msg("Creating differential expression heatmap...")

plot_difexp_heatmap(
  exprs = exprs_corrected,
  difexp = difexp_filtered,
  pdata = pdata_aligned,
  top_n = config$visualization$heatmap_top_genes,
  output_file = file.path(output_dirs$plots, "difexp_heatmap.png")
)
```

---

# 6. Network Clustering

Build protein-protein interaction network and detect communities using fastgreedy algorithm.

```{r stringdb_init}
log_msg("Initializing STRING database...")

# Initialize STRING database
string_db <- initialize_stringdb(
  version = config$stringdb$version,
  species = config$stringdb$species,
  score_threshold = config$stringdb$score_threshold,
  use_cache = config$stringdb$use_cache,
  cache_dir = config$stringdb$cache_dir
)
```

```{r map_stringdb}
log_msg("Mapping genes to STRING database...")

# Map genes to STRING
mapped <- map_to_stringdb(
  difexp = difexp_filtered,
  background_exprs = exprs_corrected,
  string_db = string_db
)

difexp_mapped <- mapped$difexp
```

```{r build_network}
log_msg("Building protein-protein interaction network...")

# Build PPI network
G <- build_ppi_network(
  string_ids = difexp_mapped$STRING_id,
  string_db = string_db
)

# Perform fastgreedy clustering
fgreedy <- fastgreedy_clustering(G)

# Add cluster assignments
difexp_clustered <- add_cluster_column(
  difexp = difexp_mapped,
  fgreedy = fgreedy,
  G = G
)

# Save clustered results
save_data(difexp_clustered,
          file.path(output_dirs$clusters, "difexp_clustered.csv"))
```

## Network Visualization

```{r network_plot}
log_msg("Creating network visualization...")

plot_network(
  G = G,
  difexp = difexp_clustered,
  fgreedy = fgreedy,
  output_file = file.path(output_dirs$plots, "network_clusters.png"),
  layout_algo = config$visualization$network_layout
)
```

## Recursive Clustering (Optional)

```{r recursive_clustering}
if (config$clustering$recursive) {
  log_msg("Performing recursive sub-clustering...")

  cluster_recursive(
    difexp = difexp_clustered,
    string_db = string_db,
    output_dir = output_dirs$base,
    min_cluster_size = config$clustering$min_cluster_size
  )
}
```

---

# 7. Functional Enrichment Analysis

Perform pathway enrichment analysis for each cluster.

```{r enrichment}
log_msg("Performing functional enrichment analysis...")

# Cluster-by-cluster enrichment
enrichment_summary <- cluster_enrichment(
  difexp = difexp_clustered,
  string_db = string_db,
  output_dir = output_dirs$base,
  max_pvalue = config$enrichment$max_pvalue
)

# Add term membership to genes
difexp_final <- add_term_membership(
  difexp = difexp_clustered,
  enrichment = string_enrichment(
    difexp_clustered$STRING_id,
    string_db,
    max_pvalue = config$enrichment$max_pvalue
  )
)

# Save final results
save_data(difexp_final,
          file.path(output_dirs$base, "difexp_final.csv"))
```

## Enrichment Visualizations

```{r enrichment_plots}
log_msg("Creating enrichment visualizations...")

# Cluster sizes
plot_cluster_sizes(
  summary_data = enrichment_summary,
  output_file = file.path(output_dirs$plots, "cluster_sizes.png")
)

# Enrichment coverage
plot_enrichment_coverage(
  summary_data = enrichment_summary,
  output_file = file.path(output_dirs$plots, "enrichment_coverage.png")
)
```

---

# 8. Summary and Results

```{r summary}
log_msg("Generating pipeline summary...")

# Collect summary statistics
summary_stats <- list(
  total_genes = nrow(merged_exprs),
  total_samples = ncol(merged_exprs),
  filtered_samples = ncol(exprs_corrected),
  difexp_genes_all = nrow(difexp),
  difexp_genes_filtered = nrow(difexp_filtered),
  genes_on_network = sum(difexp_clustered$on_graph),
  num_clusters = length(unique(difexp_clustered$cluster[!is.na(difexp_clustered$cluster)])),
  avg_cluster_size = mean(table(difexp_clustered$cluster)),
  total_enrichment_terms = sum(enrichment_summary$enrichment_terms),
  avg_coverage = mean(enrichment_summary$coverage)
)

# Write summary
write_summary(
  stats = summary_stats,
  output_file = file.path(output_dirs$reports, "pipeline_summary.csv")
)

# Print summary
cat("\n=== Pipeline Summary ===\n")
cat("Total genes analyzed:", summary_stats$total_genes, "\n")
cat("Total samples:", summary_stats$total_samples, "\n")
cat("Filtered samples:", summary_stats$filtered_samples, "\n")
cat("\nDifferential Expression:\n")
cat("  All genes:", summary_stats$difexp_genes_all, "\n")
cat("  Filtered genes:", summary_stats$difexp_genes_filtered, "\n")
cat("\nNetwork Clustering:\n")
cat("  Genes on network:", summary_stats$genes_on_network, "\n")
cat("  Number of clusters:", summary_stats$num_clusters, "\n")
cat("  Average cluster size:", round(summary_stats$avg_cluster_size, 1), "\n")
cat("\nEnrichment:\n")
cat("  Total enrichment terms:", summary_stats$total_enrichment_terms, "\n")
cat("  Average coverage:", round(summary_stats$avg_coverage * 100, 1), "%\n")
cat("========================\n\n")
```

## Session Information

```{r session_info}
print_session_info()
```

---

# Outputs

All results have been saved to: **`r output_dirs$base`**

- **Corrected expression data**: `exprs_corrected.tsv`
- **Phenodata**: `pdata_aligned.tsv`
- **Differential expression results**: `difexp/`
- **Network clusters**: `clusters/`
- **Enrichment results**: `clusters/*/cluster_enrichment.csv`
- **Visualizations**: `plots/`
- **Summary report**: `reports/pipeline_summary.csv`

For cluster-specific enrichment results, see the `clusters/` directory where each cluster has its own subdirectory with:
- `cluster_genes.csv` - Genes in the cluster
- `cluster_enrichment.csv` - Enrichment results
- `cluster_to_revigo.tsv` - Input for REVIGO analysis

---

**Pipeline completed successfully!** `r Sys.time()`
